\textbf{Evaluation of machine learning methods for 16S rRNA gene data}

### 2019-05-16 submission/manuscript.Rmd
update'

We used a cohort of 490 patients with 261 cases of SRN. For each patient fecal sample, we had `r paste(ncol(data)-1)` features (`r paste(ncol(data)-1)` OTUs) and a two-class label that defines the colonic health of the patients (SRN or normal). All the cases were independently labeled through colonoscopies. We established modeling pipelines for a binary prediction task with L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost to emphasize the differences in  model accuracy, complexity, interpretibility and computational efficiency due to  model selection. We randomly split the data into training/validation and test sets so that the training/validation set consisted of 80% of the full dataset while the test set was composed of the remaining data [Figure 1]. Since the cases are not uniformly represented in the data, the initial data-split was stratified to maintain the overall label distribution on the training set. Training/validation set consisted of 393 patients (209 SRN), while the test set was composed of 97 patients (52 SRN). The training/validation data was used for training purposes and validation of hyperparameter selection, and the test set was used for evaluation purposes. Validation of hyperparameter selection was performed using 100 randomizations of five-fold cross-validation on the training/validation set [Figure 1]. Similar to the initial data-split, five-fold cross-validation was also stratified to maintain the overall label distribution on the training and validation sets. We validated the cross-validation performances of each hyperparameter setting over the 100 randomizations and selected the best performing hyper-parameter setting to train the full training/validation dataset [Figures S1 and S2]. We then used the held-out test set to evaluate the prediction performance of each ML model. The data-split, hyperparameter selection, training and testing steps were repeated 100 times to get a reliable and robust reading of model prediction performance [Figure 1].

__The prediction and generalization performance of classifiers during cross-validation and when applied to the held-out test data.__

We evaluated the prediction performance of seven binary classification models when applied to held-out test data over 100 data-splits using the area under the receiver operating characteristic curve (AUROC) as the discriminative performance metric. Random Forest had the highest mean AUROC for detecting SRN, `r paste(round(mean((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="Random_Forest"))$test_aucs), 3))`) [Figure 2]. L2 logistic regression and XGBoost followed random forest however they had significantly lower AUROC values, `r paste(round(mean((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))`) and `r paste(round(mean((all %>% filter(model=="XGBoost"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="XGBoost"))$test_aucs), 3))`) [Figure 2]. Random forest also had significanly higher performances than L2 linear SVM, RBF SVM and decision tree which had mean AUROC values of `r paste(round(mean((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))`),  `r paste(round(mean((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))`), and `r paste(round(mean((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))`), respectively [Figure 2]. We also evaluated the generalization performance of each classifier by comparing their mean cross-validation AUROC and mean testing AUROC. We found that mean cross-validation and testing AUROC difference for each model was below 0.01.

__The complexity and interpretibility of each classifier.__

The ML models we built using L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost increase in complexity as they decrase in interpretibility, respectively. We interpreted L1 and L2 SVM with linear kernel using the feature weights and L2 logistic regression using regression coefficients of the trained models. We calculated the mean weights and coefficients of all the features over the 100 data-splits. In the three linear models, OTUs that had the largest mean weights and drove the detection of SRNs belong to family *Lachnospiraceae*,  and *Ruminococcaceae* (OTU01239, OTU00659, OTU00742, OTU00012, OTU00050, OTU00015, OTU00768, OTU00822, OTU00609, OTU01212, OTU00629) and genera *Gamella*  (OTU00426) [Figure 3]. We explained the feature importances in non-linear models using permutation importance on the held-out test data. For the tree-based models, permuting *Peptostreptococcus* (OTU00367) abundances randomly, dropped the predictive performances the most. Decision tree, random forest and XGBoost models' predictive performance dropped from `r paste(round(dt_median, 2))` base testing AUROC median to `r paste(round(median((dt_imp %>% filter(names=="Otu00367"))$new_auc), 2))`, from `r paste(round(rf_median, 2))` to `r paste(round(median((rf_imp %>% filter(names=="Otu00367"))$new_auc), 2))` and from `r paste(round(xgboost_median, 2))` to `r paste(round(median((xgboost_imp %>% filter(names=="Otu00367"))$new_auc), 2))`, respectively [Figure 4]. The negative predictive impact of *Peptostreptococcus* in the decision tree model was followed by a *Lachnospiraceae* species (OTU00058) (`r paste(round(dt_median, 2))` base testing AUROC median to `r paste(round(median((dt_imp %>% filter(names=="Otu00058"))$new_auc), 2))`) [Figure 4B]. Other OTUs had none to minimal effect on the predictive performance.

__The computational efficiency of each classifier.__

@@

#### Data collection and study population
  The data used for this analysis are stool bacterial abundances and clinical information of the patients recruited by Great Lakes-New England Early Detection Research Network study. These data were obtained from Sze et al [@sze_leveraging_2018]. 
  The stool samples were provided by recruited adult participants who were undergoing scheduled screening or surveillance colonoscopy. Colonoscopies were performed and fecal samples were collected from participants in four locations: Toronto (ON, Canada), Boston (MA, USA), Houston (TX, USA), and Ann Arbor (MI, USA). Patients' colonic disease status was defined by colonoscopy with adequate preparation and tissue histopathology of all resected lesions. Patients with an adenoma greater than 1 cm, more than three adenomas of any size, or an adenoma with villous histology were classified as advanced adenoma. Study had `r paste(sum(meta$dx=="normal"))` patients with normal colonoscopies, `r paste(sum(meta$dx=="adenoma"))` with adenomas and `r paste(sum(meta$dx=="cancer"))` with carcinomas. Of the `r paste(sum(meta$dx=="adenoma"))` adenomas, `r paste(sum(meta$Dx_Bin=="adv Adenoma"))` were identified as advanced adenomas. Stool provided by the patients was used for 16S rRNA gene sequencing to measure bacterial population abundances. The bacterial abundance data was generated by Sze et al, by processing 16S rRNA sequences in Mothur (v1.39.3) using the default quality filtering methods, identifying and removing chimeric sequences using VSEARCH and assigning to OTUs at 97% similarity using the OptiClust algorithm [@schloss_introducing_2009; @westcott_opticlust_2017; @rognes_vsearch_2016].

@@

\newpage
\includegraphics{Figure_1}
**Figure 1. Machine learning pipeline showing predictive model training and evaluation flowchart.  **  We split the data 80%/20% stratified to maintain the overall label distribution, performed five-fold cross-validation on the training data to select the best hyperparameter setting and then using these hyperparameters to train all of the training data. The model was evaluated on a held-out set of data (not used in selecting the model).
Abbreviations: AUROC, area under the receiver operating characteristic curve
\newpage
\includegraphics{Figure_2.png}
**Figure 2. Generalization and classification performance of ML models using AUROC values of all cross validation and testing performances. **   The median AUROC for diagnosing individuals with SRN using bacterial abundances was higher than chance (depicted by horizontal line at 0.50) for all the ML models. Discriminative performance of random forest model was higher than other ML models. The boxplot shows quartiles at the box ends and the statistical median as the horizontal line in the box. The whiskers show the farthest points that are not outliers. Outliers are data points that are not within 3/2 times the interquartile ranges. 
Abbreviations: SRN, screen-relevant neoplasias; AUROC, area under the receiver operating characteristic curve; SVM, support vector machine; XGBoost, extreme gradient boosting
\newpage
\includegraphics{Figure_3.png}
**Figure 3. Interpretation of the linear ML models.** (A) L2 logistic regression coefficients  (B) L1 SVM with linear kernel feature weights (C) L2 SVM with linear kernel feature weights. The means weights and coefficients of the most important five OTUs for each model are shown here with the standard deviation over 100 data-splits. Similar OTUs had the largest impact on the predictive performance of L2 logistic regression and L2 SVM with linear kernel. 
Abbreviations: SVM, support vector machine; OTU, Operational Taxonomic Unit.
\newpage
\includegraphics{Figure_4.png}
**Figure 4. Explanation of the non-linear ML models.** (A) SVM with radial basis kernel (B) decision tree (C) random forest (D) XGboost feaure importances were explanied using permutation importance using held-out test set. For all the tree-based models, a **Peptostreptococcus** species (OTU00367) had the largest impact on predictive performance of the model. 
Abbreviations: SVM, support vector machine; OTU, Operational Taxonomic Unit; RBF, radial basis kernel; OTU, Operational Taxonomic Unit.
\newpage
\includegraphics{Figure_5.png}
**Figure 5. Computational efficiency of seven ML models.** The walltimes for training and testing of each data-split showed the differences in computational efficieny of the seven models. The median walltime in hours was the highest for XGBoost and shortest for L2 logistic regression. The boxplot shows quartiles at the box ends and the statistical median as the horizontal line in the box. The whiskers show the farthest points that are not outliers. Outliers are data points that are not within 3/2 times the interquartile ranges. 
Abbreviations: AUROC, area under the receiver operating characteristic curve; SVM, support vector machine; XGBoost, extreme gradient boosting.  
\newpage

>>>>

We used a cohort of 490 patients with 261 cases of SRN. For each patient, we had `r paste(ncol(data)-1)` features (fecal bacterial abundances) and a two-class label that defines their colonic health (SRN or normal). All the cases were independently labeled through colonoscopies. We established modeling pipelines for a binary prediction task with L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost to emphasize the differences in  model accuracy, complexity, interpretibility and computational efficiency due to  model selection. We randomly split the data into training/validation and test sets so that the training/validation set consisted of 80% of the full dataset while the test set was composed of the remaining data [Figure 1]. Since the cases are not uniformly represented in the data, the initial data-split was stratified to maintain the overall label distribution on both the training/validation and test sets. Training/validation set consisted of 393 patients (209 SRN), while the test set was composed of 97 patients (52 SRN). The training/validation data was used for training purposes and validation of hyperparameter selection, and the test set was used for evaluation purposes. Validation of hyperparameter selection was performed using 100 randomizations of five-fold cross-validation on the training/validation set [Figure 1]. Similar to the initial data-split, five-fold cross-validation was also stratified to maintain the overall label distribution on the training and validation sets. We validated the cross-validation performances of each hyperparameter setting over the 100 randomizations and selected the best performing hyper-parameter setting to train the full training/validation dataset [Figures S1 and S2]. We then used the held-out test set to evaluate the prediction performance of each ML model. The data-split, hyperparameter selection, training and testing steps were repeated 100 times to get a reliable and robust reading of model prediction performance [Figure 1].

__The prediction and generalization performance of seven ML models during cross-validation and when applied to the held-out test data.__

We evaluated the prediction performance of seven binary classification models when applied to held-out test data over 100 data-splits using the area under the receiver operating characteristic curve (AUROC) as the discriminative performance metric. Random forest had the highest mean AUROC for detecting SRN, `r paste(round(mean((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="Random_Forest"))$test_aucs), 3))`), followed by L2 logistic regression and XGBoost with mean AUROC values of `r paste(round(mean((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))`) and `r paste(round(mean((all %>% filter(model=="XGBoost"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="XGBoost"))$test_aucs), 3))`) respectively [Figure 2]. Random forest had significanly higher predictive performances than L2 SVM with linear and radial basis function kernels and decision tree which had mean AUROC values of `r paste(round(mean((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))`),  `r paste(round(mean((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))`), and `r paste(round(mean((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))`), respectively [Figure 2]. We also evaluated the generalization performance of each ML model by comparing their mean cross-validation AUROC and mean testing AUROC. We found that mean cross-validation and testing AUROC difference for each model was below 0.01.

__The complexity and interpretibility of each ML model.__

The ML models we built using L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost increase in complexity as they decrase in interpretibility. We interpreted L1 and L2 SVM with linear kernel using the feature weights and L2 logistic regression using regression coefficients of the trained models. We calculated the mean weights and coefficients of all the features over the 100 data-splits. In the three linear models, OTUs that had the largest mean weights and drove the detection of SRNs belong to family *Lachnospiraceae*,  and *Ruminococcaceae* (OTU01239, OTU00659, OTU00742, OTU00012, OTU00050, OTU00015, OTU00768, OTU00822, OTU00609, OTU01212, OTU00629) and genera *Gamella*  (OTU00426) [Figure 3]. We explained the feature importances in non-linear models using permutation importance on the held-out test data. For the tree-based models, permuting *Peptostreptococcus* (OTU00367) abundances randomly, dropped the predictive performances the most. Decision tree, random forest and XGBoost models' predictive performance dropped from `r paste(round(dt_median, 2))` base testing AUROC median to `r paste(round(median((dt_imp %>% filter(names=="Otu00367"))$new_auc), 2))`, from `r paste(round(rf_median, 2))` to `r paste(round(median((rf_imp %>% filter(names=="Otu00367"))$new_auc), 2))` and from `r paste(round(xgboost_median, 2))` to `r paste(round(median((xgboost_imp %>% filter(names=="Otu00367"))$new_auc), 2))`, respectively [Figure 4]. The negative predictive impact of *Peptostreptococcus* in the decision tree model was followed by a *Lachnospiraceae* species (OTU00058) (`r paste(round(dt_median, 2))` base testing AUROC median to `r paste(round(median((dt_imp %>% filter(names=="Otu00058"))$new_auc), 2))`) [Figure 4B]. Other OTUs had none to minimal effect on the predictive performance.

__The computational efficiency of each ML model.__

@@

#### Data collection and study population
  The data used for this analysis are stool bacterial abundances and clinical information of the patients recruited by Great Lakes-New England Early Detection Research Network study. These data were obtained from Sze et al [@sze_leveraging_2018]. 
  The stool samples were provided by recruited adult participants who were undergoing scheduled screening or surveillance colonoscopy. Colonoscopies were performed and fecal samples were collected from participants in four locations: Toronto (ON, Canada), Boston (MA, USA), Houston (TX, USA), and Ann Arbor (MI, USA). Patients' colonic health was labeled by colonoscopy with adequate preparation and tissue histopathology of all resected lesions. Patients with an adenoma greater than 1 cm, more than three adenomas of any size, or an adenoma with villous histology were classified as advanced adenoma. Study had `r paste(sum(meta$dx=="normal"))` patients with normal colonoscopies, `r paste(sum(meta$dx=="adenoma"))` with adenomas and `r paste(sum(meta$dx=="cancer"))` with carcinomas. Of the `r paste(sum(meta$dx=="adenoma"))` adenomas, `r paste(sum(meta$Dx_Bin=="adv Adenoma"))` were identified as advanced adenomas. Stool provided by the patients was used for 16S rRNA gene sequencing to measure bacterial population abundances. The bacterial abundance data was generated by Sze et al, by processing 16S rRNA sequences in Mothur (v1.39.3) using the default quality filtering methods, identifying and removing chimeric sequences using VSEARCH and assigning to OTUs at 97% similarity using the OptiClust algorithm [@schloss_introducing_2009; @westcott_opticlust_2017; @rognes_vsearch_2016].

@@

\newpage
\includegraphics{Figure_1}
**Figure 1. Machine learning pipeline showing predictive model training and evaluation flowchart.  **  We split the data 80%/20% stratified to maintain the overall label distribution, performed five-fold cross-validation on the training data to select the best hyperparameter setting and then using these hyperparameters to train all of the training data. The model was evaluated on a held-out set of data (not used in selecting the model).
Abbreviations: AUROC, area under the receiver operating characteristic curve
\newpage
\includegraphics{Figure_2.png}
**Figure 2. Generalization and classification performance of ML models using AUROC values of all cross validation and testing performances. **   The median AUROC for diagnosing individuals with SRN using bacterial abundances was higher than chance (depicted by horizontal line at 0.50) for all the ML models. Discriminative performance of random forest model was higher than other ML models. The boxplot shows quartiles at the box ends and the statistical median as the horizontal line in the box. The whiskers show the farthest points that are not outliers. Outliers are data points that are not within 3/2 times the interquartile ranges. 
Abbreviations: SRN, screen-relevant neoplasias; AUROC, area under the receiver operating characteristic curve; SVM, support vector machine; XGBoost, extreme gradient boosting
\newpage
\includegraphics{Figure_3.png}
**Figure 3. Interpretation of the linear ML models.** (A) L2 logistic regression coefficients  (B) L1 SVM with linear kernel feature weights (C) L2 SVM with linear kernel feature weights. The means weights and coefficients of the most important five OTUs for each model are shown here with the standard deviation over 100 data-splits. Similar OTUs had the largest impact on the predictive performance of L2 logistic regression and L2 SVM with linear kernel. 
Abbreviations: SVM, support vector machine; OTU, Operational Taxonomic Unit.
\newpage
\includegraphics{Figure_4.png}
**Figure 4. Explanation of the non-linear ML models.** (A) SVM with radial basis kernel (B) decision tree (C) random forest (D) XGboost feaure importances were explanied using permutation importance using held-out test set. For all the tree-based models, a *Peptostreptococcus* species (OTU00367) had the largest impact on predictive performance of the model. 
Abbreviations: SVM, support vector machine; OTU, Operational Taxonomic Unit; RBF, radial basis kernel; OTU, Operational Taxonomic Unit.
\newpage
\includegraphics{Figure_5.png}
**Figure 5. Computational efficiency of seven ML models.** The walltimes for training and testing of each data-split showed the differences in computational efficieny of the seven models. The median walltime in hours was the highest for XGBoost and shortest for L2 logistic regression. The boxplot shows quartiles at the box ends and the statistical median as the horizontal line in the box. The whiskers show the farthest points that are not outliers. Outliers are data points that are not within 3/2 times the interquartile ranges. 
Abbreviations: AUROC, area under the receiver operating characteristic curve; SVM, support vector machine; XGBoost, extreme gradient boosting.  
\newpage

### 2019-05-16 submission/manuscript.tex
update'

We used a cohort of 490 patients with 261 cases of SRN. For each patient
fecal sample, we had 6920 features (6920 OTUs) and a two-class label
that defines the colonic health of the patients (SRN or normal). All the
cases were independently labeled through colonoscopies. We established
modeling pipelines for a binary prediction task with L2-regularized
logistic regression, L1 and L2 support vector machines (SVM) with linear
and radial basis function kernels, a decision tree, random forest and
XGBoost to emphasize the differences in model accuracy, complexity,
interpretibility and computational efficiency due to model selection. We
randomly split the data into training/validation and test sets so that
the training/validation set consisted of 80\% of the full dataset while
the test set was composed of the remaining data {[}Figure 1{]}. Since
the cases are not uniformly represented in the data, the initial
data-split was stratified to maintain the overall label distribution on
the training set. Training/validation set consisted of 393 patients (209
SRN), while the test set was composed of 97 patients (52 SRN). The
training/validation data was used for training purposes and validation
of hyperparameter selection, and the test set was used for evaluation
purposes. Validation of hyperparameter selection was performed using 100
randomizations of five-fold cross-validation on the training/validation
set {[}Figure 1{]}. Similar to the initial data-split, five-fold
cross-validation was also stratified to maintain the overall label
distribution on the training and validation sets. We validated the
cross-validation performances of each hyperparameter setting over the
100 randomizations and selected the best performing hyper-parameter
setting to train the full training/validation dataset {[}Figures S1 and
S2{]}. We then used the held-out test set to evaluate the prediction
performance of each ML model. The data-split, hyperparameter selection,
training and testing steps were repeated 100 times to get a reliable and
robust reading of model prediction performance {[}Figure 1{]}.

\textbf{The prediction and generalization performance of classifiers
during cross-validation and when applied to the held-out test data.}

We evaluated the prediction performance of seven binary classification
models when applied to held-out test data over 100 data-splits using the
area under the receiver operating characteristic curve (AUROC) as the
discriminative performance metric. Random Forest had the highest mean
AUROC for detecting SRN, 0.697 (std ± 0.037) {[}Figure 2{]}. L2 logistic
regression and XGBoost followed random forest however they had
significantly lower AUROC values, 0.676 (std ± 0.042) and 0.678 (std ±
0.04) {[}Figure 2{]}. Random forest also had significanly higher
performances than L2 linear SVM, RBF SVM and decision tree which had
mean AUROC values of 0.671 (std ± 0.045), 0.663 (std ± 0.044), and 0.598
(std ± 0.044), respectively {[}Figure 2{]}. We also evaluated the
generalization performance of each classifier by comparing their mean
cross-validation AUROC and mean testing AUROC. We found that mean
cross-validation and testing AUROC difference for each model was below
0.01.

\textbf{The complexity and interpretibility of each classifier.}

@@

\textbf{The computational efficiency of each classifier.}

@@

\includegraphics{Figure_1} \textbf{Figure 1. Machine learning pipeline
showing predictive model training and evaluation flowchart. } We split
the data 80\%/20\% stratified to maintain the overall label
distribution, performed five-fold cross-validation on the training data
to select the best hyperparameter setting and then using these
hyperparameters to train all of the training data. The model was
evaluated on a held-out set of data (not used in selecting the model).
Abbreviations: AUROC, area under the receiver operating characteristic
curve \newpage
\includegraphics{Figure_2.png} \textbf{Figure 2. Generalization and
classification performance of ML models using AUROC values of all cross
validation and testing performances. } The median AUROC for diagnosing
individuals with SRN using bacterial abundances was higher than chance
(depicted by horizontal line at 0.50) for all the ML models.
Discriminative performance of random forest model was higher than other
ML models. The boxplot shows quartiles at the box ends and the
statistical median as the horizontal line in the box. The whiskers show
the farthest points that are not outliers. Outliers are data points that
are not within 3/2 times the interquartile ranges. Abbreviations: SRN,
screen-relevant neoplasias; AUROC, area under the receiver operating
characteristic curve; SVM, support vector machine; XGBoost, extreme
gradient boosting \newpage
\includegraphics{Figure_3.png} \textbf{Figure 3. Interpretation of the
linear ML models.} (A) L2 logistic regression coefficients (B) L1 SVM
with linear kernel feature weights (C) L2 SVM with linear kernel feature
weights. The means weights and coefficients of the most important five
OTUs for each model are shown here with the standard deviation over 100
data-splits. Similar OTUs had the largest impact on the predictive
performance of L2 logistic regression and L2 SVM with linear kernel.
Abbreviations: SVM, support vector machine; OTU, Operational Taxonomic
Unit. \newpage
\includegraphics{Figure_4.png} \textbf{Figure 4. Explanation of the
non-linear ML models.} (A) SVM with radial basis kernel (B) decision
tree (C) random forest (D) XGboost feaure importances were explanied
using permutation importance using held-out test set. For all the
tree-based models, a \textbf{Peptostreptococcus} species (OTU00367) had
the largest impact on predictive performance of the model.
Abbreviations: SVM, support vector machine; OTU, Operational Taxonomic
Unit; RBF, radial basis kernel; OTU, Operational Taxonomic Unit.
\newpage
\includegraphics{Figure_5.png} \textbf{Figure 5. Computational
efficiency of seven ML models.} The walltimes for training and testing
of each data-split showed the differences in computational efficieny of
the seven models. The median walltime in hours was the highest for
XGBoost and shortest for L2 logistic regression. The boxplot shows
quartiles at the box ends and the statistical median as the horizontal
line in the box. The whiskers show the farthest points that are not
outliers. Outliers are data points that are not within 3/2 times the
interquartile ranges. Abbreviations: AUROC, area under the receiver
operating characteristic curve; SVM, support vector machine; XGBoost,
extreme gradient boosting.\\
\newpage

>>>>

We used a cohort of 490 patients with 261 cases of SRN. For each
patient, we had 6920 features (fecal bacterial abundances) and a
two-class label that defines their colonic health (SRN or normal). All
the cases were independently labeled through colonoscopies. We
established modeling pipelines for a binary prediction task with
L2-regularized logistic regression, L1 and L2 support vector machines
(SVM) with linear and radial basis function kernels, a decision tree,
random forest and XGBoost to emphasize the differences in model
accuracy, complexity, interpretibility and computational efficiency due
to model selection. We randomly split the data into training/validation
and test sets so that the training/validation set consisted of 80\% of
the full dataset while the test set was composed of the remaining data
{[}Figure 1{]}. Since the cases are not uniformly represented in the
data, the initial data-split was stratified to maintain the overall
label distribution on both the training/validation and test sets.
Training/validation set consisted of 393 patients (209 SRN), while the
test set was composed of 97 patients (52 SRN). The training/validation
data was used for training purposes and validation of hyperparameter
selection, and the test set was used for evaluation purposes. Validation
of hyperparameter selection was performed using 100 randomizations of
five-fold cross-validation on the training/validation set {[}Figure
1{]}. Similar to the initial data-split, five-fold cross-validation was
also stratified to maintain the overall label distribution on the
training and validation sets. We validated the cross-validation
performances of each hyperparameter setting over the 100 randomizations
and selected the best performing hyper-parameter setting to train the
full training/validation dataset {[}Figures S1 and S2{]}. We then used
the held-out test set to evaluate the prediction performance of each ML
model. The data-split, hyperparameter selection, training and testing
steps were repeated 100 times to get a reliable and robust reading of
model prediction performance {[}Figure 1{]}.

\textbf{The prediction and generalization performance of seven ML models
during cross-validation and when applied to the held-out test data.}

We evaluated the prediction performance of seven binary classification
models when applied to held-out test data over 100 data-splits using the
area under the receiver operating characteristic curve (AUROC) as the
discriminative performance metric. Random forest had the highest mean
AUROC for detecting SRN, 0.697 (std ± 0.037) followed by L2 logistic
regression and XGBoost with mean AUROC values of 0.676 (std ± 0.042) and
0.678 (std ± 0.04) respectively {[}Figure 2{]}. Random forest had
significanly higher predictive performances than those as well as L2
linear SVM, RBF SVM and decision tree which had mean AUROC values of
0.671 (std ± 0.045), 0.663 (std ± 0.044), and 0.598 (std ± 0.044),
respectively {[}Figure 2{]}. We also evaluated the generalization
performance of each ML model by comparing their mean cross-validation
AUROC and mean testing AUROC. We found that mean cross-validation and
testing AUROC difference for each model was below 0.01.

\textbf{The complexity and interpretibility of each ML model.}

@@

\textbf{The computational efficiency of each ML model.}

@@

\includegraphics{Figure_1} \textbf{Figure 1. Machine learning pipeline
showing predictive model training and evaluation flowchart. } We split
the data 80\%/20\% stratified to maintain the overall label
distribution, performed five-fold cross-validation on the training data
to select the best hyperparameter setting and then using these
hyperparameters to train all of the training data. The model was
evaluated on a held-out set of data (not used in selecting the model).
Abbreviations: AUROC, area under the receiver operating characteristic
curve \newpage
\includegraphics{Figure_2.png} \textbf{Figure 2. Generalization and
classification performance of ML models using AUROC values of all cross
validation and testing performances. } The median AUROC for diagnosing
individuals with SRN using bacterial abundances was higher than chance
(depicted by horizontal line at 0.50) for all the ML models.
Discriminative performance of random forest model was higher than other
ML models. The boxplot shows quartiles at the box ends and the
statistical median as the horizontal line in the box. The whiskers show
the farthest points that are not outliers. Outliers are data points that
are not within 3/2 times the interquartile ranges. Abbreviations: SRN,
screen-relevant neoplasias; AUROC, area under the receiver operating
characteristic curve; SVM, support vector machine; XGBoost, extreme
gradient boosting \newpage
\includegraphics{Figure_3.png} \textbf{Figure 3. Interpretation of the
linear ML models.} (A) L2 logistic regression coefficients (B) L1 SVM
with linear kernel feature weights (C) L2 SVM with linear kernel feature
weights. The means weights and coefficients of the most important five
OTUs for each model are shown here with the standard deviation over 100
data-splits. Similar OTUs had the largest impact on the predictive
performance of L2 logistic regression and L2 SVM with linear kernel.
Abbreviations: SVM, support vector machine; OTU, Operational Taxonomic
Unit. \newpage
\includegraphics{Figure_4.png} \textbf{Figure 4. Explanation of the
non-linear ML models.} (A) SVM with radial basis kernel (B) decision
tree (C) random forest (D) XGboost feaure importances were explanied
using permutation importance using held-out test set. For all the
tree-based models, a \emph{Peptostreptococcus} species (OTU00367) had
the largest impact on predictive performance of the model.
Abbreviations: SVM, support vector machine; OTU, Operational Taxonomic
Unit; RBF, radial basis kernel; OTU, Operational Taxonomic Unit.
\newpage
\includegraphics{Figure_5.png} \textbf{Figure 5. Computational
efficiency of seven ML models.} The walltimes for training and testing
of each data-split showed the differences in computational efficieny of
the seven models. The median walltime in hours was the highest for
XGBoost and shortest for L2 logistic regression. The boxplot shows
quartiles at the box ends and the statistical median as the horizontal
line in the box. The whiskers show the farthest points that are not
outliers. Outliers are data points that are not within 3/2 times the
interquartile ranges. Abbreviations: AUROC, area under the receiver
operating characteristic curve; SVM, support vector machine; XGBoost,
extreme gradient boosting.\\
\newpage
